---
title: "Multiple regression and smoothing splines for a Bayesian model: predicting daily ozone levels"  
author: "Cimmaron Yeoman"
date: "2023-04-25"
output: pdf_document
html_document: default
header-includes:
    - \usepackage{sectsty}
    - \allsectionsfont{\color{blue}}
---

## Introduction

This data set includes the environmental variables ozone, humidity, and temperature, along
with day of the week and day of the year. Ozone has some level of a relationship with human 
disturbances, especially actions which alter temperature, such as vehicle use and and 
greenhouse gas outputs. This report will attempt to predict the daily ozone level, our
response, with a multiple regression Bayesian framework. A smoothing spline will be implemented
to help make predictions and identify patterns in the data set. 

```{r, include = FALSE, warning = FALSE, message = FALSE}
library(readr)
library(rethinking)
library(knitr)
ozone <- read_csv("ozone.csv")
d <- ozone
```

## Data organization and cleaning

The data set contains categorical and index variables including days of the year,
and days of the week. These values cannot be standardized before building models. 
I insured both of these variables were stored as integers. Ozone (**OZ**), humidity (**H**), 
and temperature (**FT**) in degrees Fahrenheit were all standardized and renamed. 

```{r, include = FALSE}
d$OZ <- standardize(d$O3)
d$H <- standardize(d$humidity)
d$FT <- standardize(d$temp)
d$DW_id <- as.integer(d$dow)
d$DY <- as.integer(d$doy)
```

## Exploring the data

After standardizing, we can see from the data set summary that the average values for ozone, 
humidity, and temperature (unsquared) were at or around zero, and 75% of the observations fell below roughly 
0.75 for each. 

```{r, echo = FALSE}
summary(d[c("OZ", "H", "FT")])
```

### Density plots of ozone, humidty and temperature

The density plots below show the distribution of the standardized variables. Ozone was
skewed right and had a long tail, but if you compare this to the boxplot of ozone, the tail 
can be explained by some outliers. Humidity looked a bit odd with a bimodal appearance,
although the shorter peak on the left was clearly much smaller than the peak on the right.
Temperature had a double peak appearance as well. This can be explained by seasonal 
differences most likely, especially temperature. 

**Figure 1**
\vspace*{-2mm}
```{r, echo = FALSE, out.width = '35%'}
dens(d$OZ, col = "blue", xlim = c(-2, 4))
dens(d$H, col = "darkgreen")
dens(d$FT, col = "red", xlim = c(-3, 3))
```

### Boxplots of ozone, humidity, and temperature 

**Figure 2**
\vspace*{-10mm}
```{r, echo = FALSE, out.width = '85%'}
boxplot(d[c("OZ", "H", "FT")], ylim = c(-4, 4), 
        col = c("blue", "darkgreen", "red"), names = c("Ozone", "Humidity", "Temperature"))
```

You can see in the scatter plots why it is necessary to standardize varibles used in
models, first. Without their varying scales, they can be compared more easily
against each other. The potential outlier is visible in the ozone boxplot. 

## Scatter plots and relationships, spline candidates

**Figure 3**
\vspace*{-5mm}
```{r, echo = FALSE, out.width = '85%'}
pairs(d[c("OZ", "H", "FT", "DW_id", "DY")], pch = 20, col = "lightseagreen", 
      lower.panel = NULL)
```

Ozone had a positive linear relationship with temperature (**FT**), and some level
of overlapping observations for humidity and days of the year. The days of the week
variable (**DW_ID**) needs to be properly indexed into any models, so we can actually
observe its relationship to ozone. Days of the week produces a concave shape when
paired with different variables. These patterns could work for building the smooths. 

## Creating smoothing splines 

I knew in advance there were no missing cases, but I stored the data set **d** as a
data frame using the complete.cases function, and renamed it **d2**. Ruling out missing cases
must be done before modelling. The final choice of knots for the smooths were decided after 
trying a few different values. Degrees was kept at 3 for cubic splines. One other version like 
the spline set up below was also created for a second smooth, but was hidden to save space. Only 
the knots changed in this version. 

```{r, results = 'hide', message = FALSE}
library(splines)
d2 <- d[complete.cases(d), ]
num_knots <- 15
knot_list <- quantile(d$DY, probs = seq(0, 1, length.out = num_knots))

```

```{r}
B <- bs(d2$DY,
        knots = knot_list[-c(1, num_knots)],
        degree = 3,
        intercept = TRUE
        )
```

```{r, include = FALSE}
num_knots2 <-  15
knot_list2 <- quantile(d$DY, probs = seq(0, 1, length.out = num_knots))

```

```{r, include = FALSE}
B2 <- bs(d2$DY,
        knots = knot_list2[-c(1, num_knots2)],
        degree = 3,
        intercept = TRUE
        )
```

## First model 

This model predicts mean ozone using an indexed days of the week variable (**DW_id**),
coefficients for humidity (**H**) and temperature (**T**), and the spline created from the days of the year 
variable (**DY**). 

```{r}
set.seed(41)
flist1 <- alist(ozone ~ dnorm(mu, sigma),
               mu <- a[DW_id] + bH*H + bFT*FT + B %*% w,
               a[DW_id] ~ dnorm(0, 1),
               bH ~ dnorm(0, 1),
               bFT ~ dnorm(0, 1),
               w ~ dnorm(0, 1),
               sigma ~ dexp(1)
              )
mod1 <- quap(flist = flist1, data = list(ozone = d2$OZ, 
                                        B = B, H = d2$H, FT = d2$FT, 
                                        DW_id = d2$DW_id), 
             start = list(w = rep(0, ncol(B)))
            )
precis(mod1, depth = 2, prob = 0.93)

```


The values for w were forced to start at 0 and the model selected the values
for the posterior. The spline matrix, **B** represents $330 \times 17$. Matrix
multiplication was performed with the %*% operator, multiplying it with $17 \times 1$ 
vector, producing an object with $330$ observations. Statistical Rethinking describes 
this as a sum linear predictor for each day of the year. 

The indexed day of the week (**DW_id**) variable provides an estimate of the ozone levels 
during different days of the week. Tuesdays and Fridays (days 2, 5) seem to have higher 
ozone levels, and Sundays (day 7) have the lowest ozone levels. 

## Mean posterior predictive intervals

This is a plot of the 93% posterior predictive interval for mean ozone (**OZ**), plotted against
day of the year (**DY**). The shade includes the smooth, the day of the week (**DW_id**) index,
humidity (**H**), and temperature (**FT**). It looks very scraggly and chaotic, although we can see 
it roughly follows the more densely clustered areas of ozone level plotted with day of the year. 

**Figure 4**
\vspace*{-10mm}
```{r, echo = FALSE, out.width = '65%'}
mu <- link( mod1 )
mu_PI <- apply(mu,2,PI,0.93)
plot( d2$DY , d2$OZ , col=col.alpha(rangi2,0.4) , pch=16,
      ylab = "Standardized ozone level", xlab = "Day of the year",
      font.lab = 2)
legend("topleft", legend = c("Prediction interval (93%)"), 
           col= col.alpha("black", 0.5), lty = 1, cex = 0.8, 
           bg = "lightcyan1", text.font = 2)
shade( mu_PI , d2$DY , col=col.alpha("black",0.5) )
```

## Thoughts on the model

The model predicting the response of ozone may potentially be too complex or over-fitting.
The interval combines many variables though which I assume is why it is more narrow as well. 
We at least get a rough idea of the pattern of ozone levels against the days of the year.
Ozone levels appear higher during warmer months as they tend to peak about halfway through
each year. 

## A simpler model and smooth

This model only uses an intercept term and the B spline variable of day of the year (**DY**)
to predict ozone. While it does not include nearly as much useful information, it managed to
produce a relatively smooth  shade that follows the distribution of ozone. The shade 
provides a wider prediction interval that is less specific, but still follows the clusteres
of ozone level measurements across each day of the year

```{r}
set.seed(41)
flistSS1 <- alist(ozone ~ dnorm(mu, sigma),
               mu <- b0 + B2 %*% w,
               b0 ~ dnorm(0, 1),
               w ~ dnorm(0, 1),
               sigma ~ dexp(1)
              )
modSS1 <- quap(flist = flistSS1, data = list(ozone = d2$OZ, 
                                        B2 = B2), 
             start = list(w = rep(0, ncol(B2)))
            )
precis(modSS1, depth = 2, prob = 0.93)

```

**Figure 5**
\vspace*{-10mm}
```{r, echo = FALSE, out.width = '65%'}
mu <- link(modSS1)
muPI <- apply(mu, 2, PI, 0.93)
plot(d2$DY, d2$OZ, col = col.alpha(rangi2, 0.5), pch = 16,
     ylab = "Standardized ozone level", xlab = "Day of the year",
     font.lab = 2)
legend("topleft", legend = c("Prediction interval (93%)"), 
           col= col.alpha("black", 0.5), lty = 1, cex = 0.8, 
           bg = "lightcyan1", text.font = 2)
shade(muPI, d2$DY, col = col.alpha("black", 0.5) )
```

\vspace*{-15mm}
## Conclusion

Using variables such as humidity, temperature, and days of the year in particular, can
create a model which predicts daily ozone levels. Using a multiple regression Bayesian model
we can create a narrow 94% prediction interval smooth combined with selected predictor variables.
When we simplify the model and decrease the number of predictors, we can make a wider 93% prediction
interval that is more general, but follows the pattern of observed ozone levels. Ozone appeared
to be seasonally influenced, with higher levels in the summer months, as ozone requires about 
17◦C to form. Humidity may be useful but appears less important than temperature or day of the
year. Ozone levels may be higher during weekdays, than weekends, possibly explained by less
traffic and commuting to work. Temperature could also increase depending on the day of the week 
in urban areas with large concentrations of vehicles. 
